Spring haul-out behaviour of seals in the Bering and Chukchi Seas: implications for abundance estimation

### 2024-08-17 analysis/paper/London_HauloutBehavior.Rmd
change title per reviewer comment
M Carter suggested this title to better capture the paper's focus on the implications of abundance estimation
reword abstract per M Carter suggestions

title: Spring haul-out behavior of seals in the Bering and Chukchi seas

@@

  forecast to decrease. Improved estimates of seal population abundance are
  needed to properly monitor the impacts of these changes over time. In this
  paper, we use hourly percent-dry data from satellite-linked bio-loggers
  deployed between 2005 and 2021 to quantify the proportion of seals hauled out
  on ice. This information is needed to accurately estimate abundance from
  aerial survey counts of ice-associated seals (i.e., to correct for the
  proportion of animals that are in the water while surveys are conducted). In
  addition to providing essential data for survey 'availability' calculations,
  our analysis also provides insights into the seasonal timing and environmental
  factors affecting haul-out behavior by ice-associated seals. We specifically
  focused on bearded (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_),
  and spotted seals (_Phoca largha_) in the Bering and Chukchi seas. Because
  ringed seals (_Phoca (pusa) hispida_) can be out of the water but hidden from view in
  snow lairs analysis of their 'availability' to surveys requires special
  consideration; therefore,they were not included in this analysis. Using
  generalized linear mixed pseudo-models to properly account for temporal
  autocorrelation, we fit models with covariates of interest (e.g., day-of-year,
  solar hour, age-sex class, wind speed, barometric pressure, temperature,
  precipitation) to examine their ability to explain variation in haul-out
  probability. We found evidence for strong diel and within-season patterns in
  haul-out behavior, as well as strong weather effects (particularly wind and
  temperature). In general, seals were more likely to haul out on ice in the
  middle of the day and when wind speed was low and temperatures were higher.
  Haul-out probability increased through March and April, peaking in May and
  early June before declining again. The timing and frequency of haul-out events
  also varied based on species and age-sex class. For ribbon and spotted seals,
  models with year effects were highly supported, indicating that the timing and
  magnitude of haul-out behavior varied among years. However, we did not find
  broad evidence that haul-out timing was linked to annual sea-ice extent. Our
  analysis emphasizes the importance of accounting for seasonal and temporal
  variation in haul-out behavior, as well as associated environmental
  covariates, when interpreting the number of seals counted in aerial surveys.

>>>>

title: Spring haul-out behaviour of seals in the Bering and Chukchi Seas: implications for abundance estimation

@@

  forecast to decrease. Robust estimates of seal population abundance are needed
  to properly monitor the impacts of these changes over time. Aerial surveys of
  seals on ice are an efficient method for counting seals but must be paired
  with estimates of the proportion of seals out of the water to derive
  population abundance. In this paper, we use hourly percent-dry data from
  satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the
  proportion of seals hauled out on ice. This information is needed to
  accurately estimate abundance from aerial survey counts of ice-associated
  seals (i.e., to correct for the proportion of animals that are in the water
  while surveys are conducted). In addition to providing essential data for
  survey 'availability' calculations, our analysis also provides insights into
  the seasonal timing and environmental factors affecting haul-out behavior by
  ice-associated seals. We specifically focused on bearded (_Erignathus
  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca
  largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa)
  hispida_) can be out of the water but hidden from view in snow lairs analysis
  of their 'availability' to surveys requires special consideration;
  therefore,they were not included in this analysis. Using generalized linear
  mixed pseudo-models to properly account for temporal autocorrelation, we fit
  models with covariates of interest (e.g., day-of-year, solar hour, age-sex
  class, wind speed, barometric pressure, temperature, precipitation) to examine
  their ability to explain variation in haul-out probability. We found evidence
  for strong diel and within-season patterns in haul-out behavior, as well as
  strong weather effects (particularly wind and temperature). In general, seals
  were more likely to haul out on ice in the middle of the day and when wind
  speed was low and temperatures were higher. Haul-out probability increased
  through March and April, peaking in May and early June before declining again.
  The timing and frequency of haul-out events also varied based on species and
  age-sex class. For ribbon and spotted seals, models with year effects were
  highly supported, indicating that the timing and magnitude of haul-out
  behavior varied among years. However, we did not find broad evidence that
  haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes
  the importance of accounting for seasonal and temporal variation in haul-out
  behavior, as well as associated environmental covariates, when interpreting
  the number of seals counted in aerial surveys.
